---
title: "Disordered *matrices* with the `disordR` package: an introduction to class `dismat`"
author: "Robin Hankin"
output: html_document
---

In this short experimental discussion I set out the `dismat` virtual
`S4` class.

The idea is that we can have a matrix-like object with row or column
indexing via `disindex` objects.  Thus class `dismat_rd_ci` has row
indexing via `dismat` objectss and column indexing via (regular) index
objects, class` dismat_ri_cd` has row indexing via index objects and
column indexing via `disindex` objects, and class `dismat_rd_cd` has
both row and column indexing via `dismat` objects.  First I will
define some extraction methods for `dismat_rd_ci` objects:

```{r defina}
M <- matrix(1:35,5,7)
a <- dismat_rd_ci(M)
a


```{r showerror1,error=TRUE}
rowSums(M) # This is INCORRECT! rowSums() should return a disord object!
colSums(M)
```

Object `a` has rows indexed by `disindex` objects and columns by
regular index objects.  Thus attempting to access rows with a regular
index object should return an error:

```{r showerror,error=TRUE}
a[1,]
```

I am not sure how this will pan out, but the only way to access rows
is to use `apply()`, for which we need a new method.  As an example,
we will extract just the rows of `a` that are even:

```{r applyapply}
(a_rowsums <- apply(a,1,sum))
(wanted <- a_rowsums%%2==0)
(e <- which(wanted))
```

So `wanted` is a boolean `disord` vector that can be used to extract
rows of `a`:

```{r wantedrowsofa}
a[wanted,]
```

But we can also extract using `e`:

```{r wantedrowsofausinge,error=TRUE}
a[e,]
```


