---
title: "Matrix multiplication with the `dismat` package"
author: "Robin Hankin"
date: "2023-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r,echo=FALSE}
names <-  c("robin","bobby","sandy")
days <- c("Mon","Tue","Wed","Thu")
things <- c("petrol","parking")
costs <- matrix(c(1.01,1.03,1.01,1.04,2.31,2.42,3.22,2.32),byrow=TRUE,2,4)
dimnames(costs) <- list(commodity=things,day=days)

wentout <- matrix(c(1,1,1, 0,1,0, 1,1,0, 0,0,1),4,3,byrow=TRUE)
dimnames(wentout) <- list(day=days,name=names)
```

We have two matrices, `costs` and `wentout`:

```{r}
costs
```


Thus on Monday, petrol cost \$1.01 and parking cost \$2.31 and on Tuesday petrol cost \$1.03, and so on.
Now `wentout`:

```{r}
wentout
```

We see that on Monday, all three people went out, on Tuesday only `bobby` went out, and so on.  We want to know
how much money each person spent on each commodity.  This is given by the standard matrix product:

```{r}
costs %*% wentout
```


So far so good. But consider this:

```{r}
costs2 <- matrix(c(1.03,1.01,1.02,1.07,2.44,2.23,3.21,2.01),byrow=TRUE,2,4)
dimnames(costs2) <- list(commodity=things,day=c("Fri","Sat","Sun","Mon"))
costs2
```

We have a matrix, `costs2` that refers to a _different_ set of days.  Now a matrix product
of `costs2` and `wentout` should really return an error because we are contracting on two
incompatible sets of days.  Matrix `costs2` refers to Fri-Mon, while `wentout` refers to `Mon-Thu`.
But:

```{r}
costs2 %*% wentout
```

We see that the matrix product operates without returning an error.  This operation should be 
prohibited, but it silently returns garbage.

Further, consider this:

```{r}
costs3 <- costs[,c(1,3,2,4)]
costs3
```

Matrix `costs3` is a perfectly meaningful costs matrix, but the days are muddled up, and the columns 
refer to Mon, Wed, Tue, Thu.   The naive matrix product `costs3 %*% wentout` should also return an 
error, but:

```{r}
costs3 %*% wentout
```

We see that the matrix product again silently returns garbage.  The `dismat` package
defines several new classes of matrices to trap this kind of undefined operation, and return a meaningful
error message.  But there are several operations that are well-defined:


```{r}
(wentout3 <- wentout[c(1,3,2,4),])
costs3 %*% wentout3
```

Above we see a perfectly defined matrix product, and it is clear that this should be allowed.
The order of the summation in the inner product is immaterial, which suggests that using a
the days as a summation variable [that is, the row index of `wentout` and the column index of `costs`]
should be a `disindex` object from the `disordR` package.

Part of the problem is that the row names of `wentout` do not have to be unique.  Consider augmenting our data
collection with a subsequent Monday:

```{r}
(costs4 <- cbind(costs,Mon=c(1.12,2.66)))
(wentout4 <- rbind(wentout,Mon=c(1,0,0)))
```

Now we have two days named `Mon`.  But this does not matter; the matrix product is well-defined:

```{r}
costs4 %*% wentout4
```


Another observation: we can dispense with the people's names and the commodity names.  We could think
about the people and indeed the commodities as being indexed by their position in the matrix:

```{r}
costs5 <- costs4 ;  rownames(costs5) <- NULL
wentout5 <- wentout4 ;  colnames(wentout5) <- NULL
costs5
costs5 %*% wentout5
```

Going back to `costs4` we could equally dispense with the names of the days:

```{r}
costs6 <- costs4; colnames(costs6) <- NULL
wentout6 <- wentout4 ; rownames(wentout6) <- NULL
costs6
wentout6
costs6 %*% wentout6
```

But in this case we also need, somehow, to check for consistency between the unnamed columns of `costs6` and the unnamed rows of `wentout6`.

